---
title: "Chi-square example"
author: "Ronnie Bailey-Steinitz"
date: "2026-07-17"
output: html_document
editor_options: 
  chunk_output_type: console
---

## 0. Load Required Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
```


1. Land use and wildlife signs

Dataset 1: Wildlife signs

**Question:** *Is the type of wildlife sign associated with land use?*

**Null hypothesis:** *If land use is not associated with the type of wildlife sign, we'd expect to see about the same proportions of sightings, dung, tracks, and feeding signs in each land-use type.*

or... if land use *doesn't* matter, then protected forests, agricultural edges, and village edges should all show a *similar* mix of wildlife signs.
```{r}
wildlife_signs <- readr::read_csv(
  here::here("Week 4/land_use_wildlife_signs.csv")
) %>%
  janitor::clean_names()

wildlife_sign_table <- table(
  wildlife_signs$land_use,
  wildlife_signs$sign_type
)

addmargins(wildlife_sign_table)

wildlife_sign_chi <- chisq.test(wildlife_sign_table)

wildlife_sign_chi
wildlife_sign_chi$expected # expected counts!
```

The chi-square statistic summarizes the total difference between the observed and expected counts across all cells in the table. The larger the X^2 value, the greater the evidence that the two categorical variables are associated.

**If the observed counts were very close to the expected counts, X^2 would be close to 0.**

Our example's value of X^2 = 22.15 tells us that, overall, **the observed counts differ quite a bit** from what we'd expect if land use and wildlife sign were unrelated.

The p-value (0.0011) tells us that differences this large would be very unlikely to occur by chance alone if there were truly no association.

| Land use          | Sign         | Observed | Expected | Interpretation                     |
| ----------------- | ------------ | -------: | -------: | ---------------------------------- |
| Protected forest  | Dung         |   **70** | **57.1** | Much more dung than expected       |
| Protected forest  | Tracks       |   **60** | **72.6** | Fewer tracks than expected         |
| Village edge      | Dung         |    **5** | **10.7** | Much less dung than expected       |
| Village edge      | Feeding sign |    **5** | **10.7** | Much less feeding than expected    |
| Village edge      | Tracks       |   **22** | **13.6** | More tracks than expected          |
| Agricultural edge | Tracks       |   **45** | **40.8** | Slightly more tracks than expected |

Meaning: 

- Protected forest: Animals spend more time there, so there's more dung than expected and relatively fewer tracks, suggesting they're not just passing through.
- Village edge: There are more tracks than expected but less dung and fewer feeding signs, suggesting animals tend to move through village edges rather than spend much time there.
- Agricultural edge: Tracks are close to (or slightly above) expectation, consistent with animals traveling through these areas to access crops or other resources.


2. Habitat and animal activity

Dataset 2: Animal activity

**Question:** *Is animal activity associated with habitat type?*

**Null hypothesis:** *If habitat type is not associated with animal activity, we'd expect to see about the same proportions of feeding, moving, resting, and socializing in each habitat.*

or... if habitat *doesn't* influence behavior, forests, grasslands, and wetlands should all have a *similar* mix of activities.
```{r}
animal_activity <- readr::read_csv(
  here::here("Week 4/habitat_animal_activity.csv")
) %>%
  janitor::clean_names()

# continue on your own!
```

**Your Turn!**

Use the `animal_activity` dataset to investigate whether habitat type is associated with animal activity.

**1. Create a contingency table**
Create a table showing the number of observations for each combination of habitat and activity.
Add row and column totals.

> Question: 

- Which habitat has the most observations? Which activity is most common overall?


**2. Run a chi-square test**

Run a chi-square test using your contingency table.

> Questions:

- What is the chi-square statistic (X^2)?
- What is the p-value?
- Based on the p-value, what do you conclude?


**3. Examine the expected counts**

Display the expected counts.

> Questions:

- If habitat type had no effect on animal activity, what pattern would you expect to see?
- Are the observed counts similar to the expected counts, or are some cells noticeably different?


**4. Compare observed and expected counts**

- Identify at least three combinations of habitat and activity that differ noticeably from the expected counts.

