---
title: "Week 4: Count, Binary, and Categorical Outcomes"
author: "Ronnie Bailey-Steinitz"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
options(width = 60)
```

# Skills Learning -- Lecture

Last week, we introduced nonparametric tests and began working with count data using a Poisson generalized linear model.

This week, we will complete several major branches of our model-selection framework:

- Poisson regression for count data
- negative binomial regression for overdispersed count data
- logistic regression for binary outcomes
- chi-square tests for relationships between categorical variables
- Fisher's exact test when categorical sample sizes are small

The main question for today is:

> **What statistical analysis should I use when my response variable is a count, binary, or categorical variable?**

## 0. Load Required Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(MASS)
```

## 1. Load and Preview the Dataset

We will continue using the faux gorilla hormone and behavioral dataset from Week 3.

```{r}
data <- readr::read_csv(
  here::here("Week 3/gorilla_hormone_data_week3.csv")
) %>%
  janitor::clean_names()

glimpse(data)
```

Create labels for binary variables so the output and figures are easier to read.

```{r}
data <- data %>%
  mutate(
    injury_label = if_else(
      injury_status == 1,
      "Injured",
      "Not injured"
    ),
    respiratory_label = if_else(
      respiratory_signs == 1,
      "Respiratory signs",
      "No respiratory signs"
    )
  )
```

# Part 1: Count Outcomes

## 2. Create a Poisson-Like Count Variable

The Week 3 dataset contains count variables, but they are strongly overdispersed. To practice a Poisson model first, we will create a new faux count variable called `feeding_bouts`.

The expected number of feeding bouts will increase with fruit availability.

```{r}
set.seed(2026)

data <- data %>%
  mutate(
    expected_feeding_bouts = exp(
      1.1 + 1.2 * fruit_availability_index
    ),
    feeding_bouts = rpois(
      n = n(),
      lambda = expected_feeding_bouts
    )
  )

glimpse(data)
summary(data$feeding_bouts)
```

The `rpois()` function generates count values from a Poisson distribution.

For each row:

- `n()` tells R how many values to generate
- `lambda` gives the expected count
- the expected count changes with fruit availability

### 2.1 Inspect the Count Variable

```{r}
mean(data$feeding_bouts)
var(data$feeding_bouts)

ggplot(data, aes(x = feeding_bouts)) +
  geom_histogram(binwidth = 1) +
  labs(
    x = "Feeding bouts",
    y = "Number of observations",
    title = "Distribution of feeding bouts"
  ) +
  theme_bw()
```

## 3. Poisson Regression

Biological question:

> **Does fruit availability predict the number of feeding bouts?**

Response variable:

- `feeding_bouts`, a count

Predictor variable:

- `fruit_availability_index`, continuous

```{r}
mod_pois <- glm(
  feeding_bouts ~ fruit_availability_index,
  data = data,
  family = "poisson"
)

summary(mod_pois)
```

### 3.1 Interpret the Coefficients

Poisson regression coefficients are on the log scale.

```{r}
coef(mod_pois)
```

Exponentiate the coefficients to interpret them on the count scale.

```{r}
exp(coef(mod_pois))
```

For a continuous predictor:

- an exponentiated coefficient greater than 1 indicates an increase
- an exponentiated coefficient less than 1 indicates a decrease
- an exponentiated coefficient equal to 1 indicates no change

Calculate the percentage change associated with a one-unit increase in fruit availability.

```{r}
fruit_rate_ratio <- exp(
  coef(mod_pois)["fruit_availability_index"]
)

fruit_percent_change <- (
  fruit_rate_ratio - 1
) * 100

fruit_rate_ratio
fruit_percent_change
```

Interpretation template:

> For each one-unit increase in fruit availability, the expected number of feeding bouts is multiplied by `fruit_rate_ratio`. This is equivalent to approximately a `fruit_percent_change`% change in expected feeding bouts.

## 4. Predicted Values from the Poisson Model

Create a sequence covering the observed range of fruit availability.

```{r}
new_fruit <- tibble(
  fruit_availability_index = seq(
    from = min(
      data$fruit_availability_index,
      na.rm = TRUE
    ),
    to = max(
      data$fruit_availability_index,
      na.rm = TRUE
    ),
    length.out = 100
  )
)
```

Use the fitted model to predict feeding bouts across this range.

```{r}
new_fruit$predicted_feeding_bouts <- predict(
  mod_pois,
  newdata = new_fruit,
  type = "response"
)

head(new_fruit)
```

Plot the observed counts and model predictions.

```{r}
ggplot(
  data,
  aes(
    x = fruit_availability_index,
    y = feeding_bouts
  )
) +
  geom_point(alpha = 0.4) +
  geom_line(
    data = new_fruit,
    aes(
      x = fruit_availability_index,
      y = predicted_feeding_bouts
    ),
    linewidth = 1
  ) +
  labs(
    x = "Fruit availability index",
    y = "Feeding bouts",
    title = "Predicted feeding bouts by fruit availability"
  ) +
  theme_bw()
```

## 5. Check for Overdispersion

A Poisson model assumes that the amount of variation is consistent with a Poisson distribution.

A quick diagnostic is:

```text
Residual deviance / residual degrees of freedom
```

```{r}
dispersion_pois <- deviance(mod_pois) /
  df.residual(mod_pois)

dispersion_pois
```

General guide:

- close to 1: little evidence of overdispersion
- greater than 1.5 or 2: possible overdispersion
- much larger than 2: strong overdispersion

This is a diagnostic guide, not a strict cutoff.

## 6. An Overdispersed Count Variable

Now return to `parasite_load`, which was designed to be much more variable than a Poisson distribution expects.

Biological question:

> **Does parasite load change with age?**

```{r}
summary(data$parasite_load)

mean(data$parasite_load)
var(data$parasite_load)

ggplot(data, aes(x = parasite_load)) +
  geom_histogram(binwidth = 5) +
  labs(
    x = "Parasite load",
    y = "Number of observations",
    title = "Distribution of parasite load"
  ) +
  theme_bw()
```

### 6.1 Fit a Poisson Model First

```{r}
mod_parasite_pois <- glm(
  parasite_load ~ age_years,
  data = data,
  family = "poisson"
)

summary(mod_parasite_pois)
```

Check overdispersion.

```{r}
dispersion_parasite <- deviance(
  mod_parasite_pois
) / df.residual(mod_parasite_pois)

dispersion_parasite
```

If this value is much larger than 1, the parasite counts vary more than the Poisson model allows.

## 7. Negative Binomial Regression

Negative binomial regression is used for count data that are overdispersed.

It follows the same formula structure as a Poisson regression but allows extra variability.

```{r}
mod_parasite_nb <- MASS::glm.nb(
  parasite_load ~ age_years,
  data = data
)

summary(mod_parasite_nb)
```

Exponentiate the coefficients.

```{r}
exp(coef(mod_parasite_nb))
```

Calculate the percentage change associated with one additional year of age.

```{r}
age_rate_ratio_nb <- exp(
  coef(mod_parasite_nb)["age_years"]
)

age_percent_change_nb <- (
  age_rate_ratio_nb - 1
) * 100

age_rate_ratio_nb
age_percent_change_nb
```

## 8. Compare Poisson and Negative Binomial Models

Use AIC to compare the two models.

```{r}
AIC(
  mod_parasite_pois,
  mod_parasite_nb
)
```

Interpretation:

> Lower AIC indicates better relative model fit. If the negative binomial model has a substantially lower AIC, it is a better choice for these overdispersed counts.

# Part 2: Binary Outcomes

## 9. What Is a Binary Outcome?

A binary variable has only two possible values.

Examples:

- injured / not injured
- sick / healthy
- present / absent
- alive / dead
- success / failure

In this dataset:

```{r}
data %>%
  count(injury_status, injury_label)

data %>%
  count(
    respiratory_signs,
    respiratory_label
  )
```

The numeric coding is:

- `0` = event did not occur
- `1` = event occurred

The mean of a 0/1 variable is the proportion of observations coded 1.

```{r}
mean(data$injury_status)
mean(data$respiratory_signs)
```

## 10. Why Not Use Linear Regression for a Binary Outcome?

Let's ask:

> **Does parasite load predict whether a gorilla is injured?**

A linear model can produce predicted values below 0 or above 1, even though probabilities must remain between 0 and 1.

```{r}
mod_binary_lm <- lm(
  injury_status ~ parasite_load,
  data = data
)

summary(mod_binary_lm)

summary(
  predict(mod_binary_lm)
)
```

Check for impossible predicted probabilities.

```{r}
data %>%
  mutate(
    predicted_lm = predict(mod_binary_lm)
  ) %>%
  filter(
    predicted_lm < 0 |
      predicted_lm > 1
  ) %>%
  select(
    sample_id,
    gorilla_id,
    parasite_load,
    injury_status,
    predicted_lm
  )
```

## 11. Logistic Regression

Logistic regression is used when the response variable is binary.

In R:

```text
glm(..., family = "binomial")
```

Biological question:

> **Does parasite load predict the probability of injury?**

```{r}
mod_logistic <- glm(
  injury_status ~ parasite_load,
  data = data,
  family = "binomial"
)

summary(mod_logistic)
```

### 11.1 Interpret Logistic Regression Coefficients

Logistic regression coefficients are on the log-odds scale.

```{r}
coef(mod_logistic)
```

Exponentiating the coefficients gives odds ratios.

```{r}
exp(coef(mod_logistic))
```

Interpretation guide:

- odds ratio greater than 1: higher predictor values are associated with higher odds of the event
- odds ratio less than 1: higher predictor values are associated with lower odds of the event
- odds ratio equal to 1: no change in odds

Calculate the odds ratio for parasite load.

```{r}
parasite_odds_ratio <- exp(
  coef(mod_logistic)["parasite_load"]
)

parasite_odds_ratio
```

Interpretation template:

> For each one-unit increase in parasite load, the odds of injury are multiplied by `parasite_odds_ratio`.

## 12. Predicted Probabilities

Predicted probabilities are usually easier to understand biologically than log-odds.

Create a sequence of parasite-load values.

```{r}
new_parasites <- tibble(
  parasite_load = seq(
    from = min(
      data$parasite_load,
      na.rm = TRUE
    ),
    to = max(
      data$parasite_load,
      na.rm = TRUE
    ),
    length.out = 100
  )
)
```

Predict the probability of injury.

```{r}
new_parasites$predicted_injury_probability <- predict(
  mod_logistic,
  newdata = new_parasites,
  type = "response"
)

head(new_parasites)
tail(new_parasites)
```

Plot the observed binary outcomes and predicted probabilities.

```{r}
ggplot(
  data,
  aes(
    x = parasite_load,
    y = injury_status
  )
) +
  geom_jitter(
    height = 0.03,
    width = 0,
    alpha = 0.35
  ) +
  geom_line(
    data = new_parasites,
    aes(
      x = parasite_load,
      y = predicted_injury_probability
    ),
    linewidth = 1
  ) +
  labs(
    x = "Parasite load",
    y = "Predicted probability of injury",
    title = "Probability of injury by parasite load"
  ) +
  theme_bw()
```

# Part 3: Categorical Variables

## 13. Contingency Tables

A contingency table displays the number of observations in each combination of two categorical variables.

Example question:

> **Is injury status associated with sex?**

```{r}
injury_sex_table <- table(
  data$sex,
  data$injury_label
)

injury_sex_table
```

Add row and column totals.

```{r}
addmargins(injury_sex_table)
```

View proportions by sex.

```{r}
prop.table(
  injury_sex_table,
  margin = 1
)
```

## 14. Chi-Square Test of Independence

A chi-square test asks whether two categorical variables are associated.

The null hypothesis is:

> Injury status and sex are independent.

The alternative hypothesis is:

> Injury status and sex are associated.

```{r}
chi_result <- chisq.test(
  injury_sex_table
)

chi_result
```

### 14.1 Expected Counts

The chi-square test compares:

- observed counts: what we actually saw
- expected counts: what we would expect if the variables were unrelated

```{r}
injury_sex_table

chi_result$expected
```

Interpretation:

- a small p-value suggests an association
- a large p-value means we do not have strong evidence of an association

Biological interpretation template:

> Injury status was / was not significantly associated with sex (chi-square test, p = ___).

## 15. Fisher's Exact Test

The chi-square approximation may be unreliable when expected cell counts are very small.

Check the expected counts:

```{r}
chi_result$expected
```

If one or more expected counts are below about 5, Fisher's exact test is often safer.

```{r}
fisher.test(
  injury_sex_table
)
```

Fisher's exact test answers the same basic question:

> Are the two categorical variables associated?

It calculates an exact p-value rather than relying on the chi-square approximation.

# Part 4: Choosing the Analysis

## 16. Summary

| Response or question | Possible analysis |
|---|---|
| Continuous response | Linear regression |
| Count response with acceptable Poisson fit | Poisson regression |
| Overdispersed count response | Negative binomial regression |
| Binary response | Logistic regression |
| Association between two categorical variables | Chi-square test |
| Two categorical variables with small expected counts | Fisher's exact test |

Before choosing an analysis, ask:

1. What is the biological question?
2. What is the response variable?
3. What type of variable is the response?
4. Are we modeling an outcome or testing an association?
5. Does the model match the structure of the response variable?
6. Are observations independent?

# Skills Application -- Lab

## 17. Lab Challenge

Work in your ongoing file:

`Lastname_Firstname_Data.Rmd`

### Task List

1. Load your dataset and continue working in your script.
2. Identify one count, binary, or categorical variable.
3. Write a biological question using that variable.
4. Choose an appropriate analysis:
   - Poisson regression
   - negative binomial regression
   - logistic regression
   - chi-square test
   - Fisher's exact test
5. Run the analysis.
6. Check any important model condition:
   - overdispersion for a Poisson model
   - expected cell counts for a chi-square test
7. Interpret the result biologically in 2-3 sentences.

> **Goal:** Match the analysis to the response variable and complete the path from research question to biological insight.

## 18. Student Template

**Biological question:**  

**Response variable:**  

**Response variable type:**  

**Predictor or second categorical variable:**  

**Chosen analysis:**  

**Why is this analysis appropriate?**  

**Important diagnostic or assumption check:**  

**Biological interpretation:**  

```{r}
# Load and inspect your data
```

```{r}
# Run your chosen analysis
```

```{r}
# Check an important model condition
```

```{r}
# Create predicted values or a plot only if it helps interpretation
```
