---
title: "Week 4: Count, Binary, and Categorical Outcomes"
author: "Ronnie Bailey-Steinitz"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r global_options, include=FALSE}
options(width = 60)
```

# Skills Learning -- Lecture

Last week, we introduced nonparametric tests and began working with count data.

This week, we will complete several major branches of our model-selection framework by asking:

> **What analysis should I use when my response variable is a count, binary, or categorical variable?**

Today we will focus on:

- negative binomial regression for count outcomes
- logistic regression for binary outcomes
- chi-square tests for relationships between categorical variables
- Fisher's exact test when expected counts are small

## 0. Load Required Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(MASS)
```

## 1. Load and Preview the Dataset

We will continue using the faux gorilla hormone and behavioral dataset.

```{r}
data <- readr::read_csv(
  here::here("Week 3/gorilla_hormone_data_week3.csv")
) %>%
  janitor::clean_names()

glimpse(data)
```

# Part 1: Count Outcomes

## 2. Count Response Variables

Count variables represent the number of times something happened.

Examples include:

- number of parasites
- number of aggression events
- number of nests
- number of vocalizations
- number of animal sightings

Count variables:

- are whole numbers
- cannot be negative
- often include many zeros
- are often highly variable in ecological datasets

Today, we will use `parasite_load` as our count response variable.

```{r}
summary(data$parasite_load)

hist(data$parasite_load)
```

## 3. Negative Binomial Regression

Negative binomial regression is used when:

- the response variable is a count
- the counts are highly variable
- the response cannot be negative

Biological question:

> **Does parasite load change with age?**

```{r}

mod_pois <- glm(parasite_load ~ age_years, 
                family = poisson,
                data = data)

deviance(mod_pois)/df.residual(mod_pois)

mod_nb <- glm.nb(parasite_load ~ age_years, data = data)

summary(mod_nb)

AIC(mod_pois, mod_nb)

exp(coef(mod_nb))

ggplot(data, aes(x = age_years, y = parasite_load)) + geom_smooth() + geom_point(alpha = 0.5)

```

> We asked: Does parasite load change with age?

Negative binomial regressions use a log link, so these coefficients are on the log scale.



... now, let's exponentiate the result:

### 3.1 Interpret the Coefficients

Negative binomial regression uses a log link, so the coefficients are initially reported on the log scale.

```{r}

# Model coefficients:
coef(mod_nb)
# Exponentiate the coefficients to interpret them on the original count scale:
exp(coef(mod_nb))
```

Interpretation guide:

- exponentiated coefficient greater than 1: expected count increases
- exponentiated coefficient less than 1: expected count decreases
- exponentiated coefficient equal to 1: no change


## 4. Predicted Values from the Negative Binomial Model

Create a sequence covering the observed age range.

```{r}

new_age <- tibble(
  age_years = seq(from = min(data$age_years, na.rm = TRUE),
                 to = 150, 
                   #max(data$age_years),
                 length.out = 100
                 ))

```

Use the fitted model to predict parasite load at each age.

```{r}

new_age$predicted_parasite_load <- predict(mod_nb, newdata = new_age, type = "response")


```

Plot the observed counts and model predictions.

```{r}

ggplot(data, aes(x = age_years, y = parasite_load)) +
  geom_point(alpha = 0.4) +
  # geom_smooth(method = "lm", fullrange = TRUE) +
  geom_line(data = new_age, aes(x = age_years, y = predicted_parasite_load)) +
  theme_bw()

```

# Part 2: Binary Outcomes

## 5. What Is a Binary Outcome?

A binary variable has only two possible values.

Examples:

- injured / not injured
- sick / healthy
- present / absent

In this dataset:

```{r}

data |>
  count(
    injury_status, 
    injury_label
  )
  

```

The coding is:

- `0` = event did not occur
- `1` = event occurred

For a variable coded 0 and 1, the mean is the proportion of observations coded 1.

```{r}

mean(data$injury_status)

```

## 6. Logistic Regression

When the **response variable is binary**, we use a **logistic regression**.

Biological question:

> **Does parasite load predict the probability of injury?**

Response variable:

- `injury_status`, binary

Predictor variable:

- `parasite_load`, count

```{r}

mod_log <- glm(injury_status ~ parasite_load,
               data = data,
               family = "binomial")

summary(mod_log)

```

This is very similar to the negative binomial output. The biggest difference is *what* we're predicting. Instead of predicting a **count**, we're predicting the **probability** that an event occurs (injured = 1, or 100%).

> Our biological question was: Does parasite load predict whether a gorilla is injured?


### 6.1 Interpret Logistic Regression Coefficients

Logistic regression coefficients are reported on the log-odds scale.

```{r}

coef(mod_log)

```

Exponentiating the coefficients gives odds ratios.

```{r}

exp(coef(mod_log))

```

Interpretation guide:

- odds ratio greater than 1: higher predictor values are associated with higher odds of the event
- odds ratio less than 1: higher predictor values are associated with lower odds of the event
- odds ratio equal to 1: no change in odds


## 7. Predicted Probabilities

Predicted probabilities are often easier to understand biologically than odds.

Create a sequence covering the observed parasite-load range.

```{r}
new_parasites <- tibble(parasite_load = seq(from = min(data$parasite_load, na.rm = TRUE),
                                            to = max(data$parasite_load, na.rm = TRUE),
                                            length.out = 100))
```

Predict the probability of injury.

```{r}

new_parasites$predicted_injury_probability <- predict(mod_log, newdata = new_parasites, type = "response")

```

Plot the observed binary outcomes and predicted probabilities.

```{r}
cutoff <- new_parasites |> 
  slice_min(abs(predicted_injury_probability - 0.5))

ggplot(data, aes(x = parasite_load, y = injury_status)) +
  geom_point(alpha = 0.4) +
  geom_line(data = new_parasites, aes(x = parasite_load, y = predicted_injury_probability)) +
  # geom_hline(yintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = cutoff$parasite_load, linetype = "dashed", color = "red") +
  theme_bw() +
  labs(x = "Parasite Load",
       y = "Injury Status",
       title = "Do parasites predict\ninjury status?")

```



# Part 3: Relationships Between Categorical Variables

## 8. Contingency Tables

A contingency table displays the number of observations in each combination of two categorical variables.

Biological question:

> **Is injury label associated with sex?**

```{r}

injury_sex_table <- table(
  data$sex,
  data$injury_label
)

injury_sex_table

prop.table(injury_sex_table, margin = 1)

```


## 9. Chi-Square Test of Independence

A chi-square test asks whether two categorical variables are associated. It first calculates **what the counts would look like if the variables were completely unrelated**, then those expected counts are **compared to the counts we actually observed**. If the observed counts are much farther from the expected counts than we would expect by random chance, the relationship is considered statistically significant.

The null hypothesis is:

> Injury status and sex are independent.

The alternative hypothesis is:

> Injury status and sex are associated.

```{r}

addmargins(injury_sex_table)

```

The chi-square test starts by assuming the null hypothesis: sex and injury are completely unrelated.

If that's true, then females and males should have the same injury rate: overall, 20 of 240 gorillas were injured.

Results:
```{r}

chi_result <- chisq.test(injury_sex_table)
chi_result

```

Interpretation:

- a small p-value suggests an association
- a large p-value means we do not have strong evidence of an association



## 10. Observed and Expected Counts

The chi-square test compares:

- **observed counts**: what we actually saw
- **expected counts**: what we would expect if the variables were unrelated

```{r}

injury_sex_table

chi_result$expected

```


> The chi-square test then asks: are these differences larger than we'd expect from random chance?

If yes -> small p-value.  
If no -> large p-value.  

Large differences between observed and expected counts contribute more strongly to the chi-square statistic.

Conceptually, it's measuring: How different are the observed counts from the expected counts?

**Small differences**
- small X^2
- large p-value

**Large differences**
- large X^2
- small p-value


## 11. Fisher's Exact Test

The chi-square approximation may be unreliable when expected counts are very small.

Check the expected counts:

```{r}
chi_result$expected
```

If one or more expected counts are below approximately 5, Fisher's exact test is often safer.

```{r}
fisher.test(
  injury_sex_table
)
```

Fisher's exact test asks the same basic question:

> Are these two categorical variables associated?

It calculates an exact p-value rather than relying on the chi-square approximation.

# Part 4: Choosing the Analysis

## 12. Summary

| Response variable or question | Possible analysis |
|---|---|
| Continuous response | Linear regression |
| Count response | Negative binomial regression |
| Binary response | Logistic regression |
| Association between two categorical variables | Chi-square test |
| Two categorical variables with small expected counts | Fisher's exact test |

Before choosing an analysis, ask:

1. What is the biological question?
2. What is the response variable?
3. What type of variable is the response?
4. Are we modeling an outcome or testing an association?
5. Does the model match the structure of the response variable?
6. Are the observations independent?

# Skills Application -- Lab

## 13. Lab Challenge

Work in your ongoing file:

`Lastname_Firstname_Data.Rmd`

### Task List

1. Load your dataset and continue working in your script.
2. Identify one count, binary, or categorical variable.
3. Write a biological question using that variable.
4. Choose an appropriate analysis:
   - negative binomial regression
   - logistic regression
   - chi-square test
   - Fisher's exact test
5. Run the analysis.
6. Interpret the result biologically in 2-3 sentences.
7. Create predicted values or a plot only if it helps explain the result.

> **Goal:** Match the analysis to the response variable and complete the path from research question to biological insight.
