---
title: "Radac_wk3"
author: "Ronnie Bailey-Steinitz"
date: "2026-07-08"
output:
  pdf_document: default
  html_document: default
editor_options: 
  chunk_output_type: console
---

# Skills Learning -- Lecture

Last week, we focused on comparing groups when the response variable was quantitative and reasonably close to normal.

This week, we will ask:

> **What do we do when normality is not a good assumption?**

This happens often in ecological and biological data. Many response variables are not normally distributed.

Examples include:

- count data, such as aggression events or parasite load
- skewed data, such as hormone concentrations
- data with many zeros
- data that cannot be negative

Today we will focus on:

- recognizing when a t-test, ANOVA, or linear model may not be appropriate
- nonparametric alternatives to t-tests and ANOVA
- count data
- Poisson regression
- overdispersion
- negative binomial regression
- when log transformations may or may not help

The main question for today is:

> **What kind of analysis should I use when my response variable is not normally distributed?**

## 0. Load Required Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
# install.packages("rstatix")
library(rstatix) # for Dunn post-hoc test
```

## 1. Load and Preview the Dataset

Today we will continue using the faux gorilla hormone + behavioral dataset.

```{r}
# load dataset
data <- readr::read_csv(here::here("Week 3/gorilla_hormone_data_week3.csv")) %>% 
  janitor::clean_names()

# preview dataset
glimpse(data)
```

## 2. When Normality Is Not Appropriate

Many standard tests assume that the response variable, or the model residuals, are approximately normal.

This is often reasonable for continuous variables like:

- body mass
- hormone concentration
- temperature
- age

But it is often not reasonable for variables like:

- number of aggression events
- parasite load
- number of nests
- number of offspring
- number of animal sightings

Count variables have special properties:

- they are whole numbers
- they cannot be negative
- they often have many zeros
- they are often right-skewed
- the variance often increases as the mean increases

### 2.1 Look at count data
```{r}
summary(data$aggression_events)

hist(data$aggression_events)
```

### Key idea

A model should match the type of response variable.

```text
Quantitative and approximately normal     ->  t-test, ANOVA, linear model
Non-normal group comparison               ->  Wilcoxon or Kruskal-Wallis
Count response                            ->  Poisson or negative binomial model
Binary response                           ->  logistic regression
Repeated measures                         ->  mixed effects model
```



## 3. Nonparametric Alternatives

Last week, we used t-tests and ANOVA to compare group means.  
But what if the response variable is strongly skewed, has extreme outliers, or does not look approximately normal?

In those cases, we may use nonparametric tests.  
Nonparametric tests are useful because they do not rely as strongly on the assumption that the response variable is normally distributed.

Today we will use:

- `wilcox.test()` as an alternative to a two-sample t-test
- `kruskal.test()` as an alternative to ANOVA

Important note:

> Nonparametric tests are not automatically better. They are used when the assumptions of the parametric test are not reasonable.


## 4. Wilcoxon Test: Alternative to a Two-Sample t-test

The Wilcoxon test is often used when we want to compare two independent groups, but the response variable is not approximately normal.

### 4.1 Visualize
```{r}

# visualize parasite load
ggplot(data, aes(x = injury_label, y = parasite_load)) +
  geom_boxplot()

data |> 
  group_by(injury_label) |> 
  # dplyr::filter(sex != "F") |> 
  summarize(
    n = n(),
    mean_parasites = mean(parasite_load, na.rm = TRUE),
    sd_parasites = sd(parasite_load, na.rm = TRUE)
  )
  
```

### 4.2 Wilcoxon Test

```{r}

wilcox.test(parasite_load ~ injury_label, data = data)

```

### How to interpret this

The Wilcoxon test asks:

> Do the values in one group tend to be higher or lower than the values in the other group?

This is slightly different from a t-test, which focuses on the difference between group means.

For non-normal or skewed data, it is often useful to report the median as well as the mean.

The Wilcoxon test doesn't compare means like a t-test does. Instead, it compares the locations of the two distributions (based on the ranks of the observations).


### 4.3 Ranking values
```{r}

ranked_data <- data |> 
  mutate(parasite_rank = rank(parasite_load))

ggplot(ranked_data, aes(x = parasite_rank, y = parasite_load, color = injury_label, shape = injury_label)) +
  geom_point(aes(alpha = injury_label), size = 2) +
  scale_alpha_manual(values = c(0.6, 0.3)) +
  stat_ellipse(level = 0.95) +
  scale_shape_manual(values = c(17, 20)) +
  theme_bw() +
  labs(x = "Rank of Parasite Load",
       y = "Parasite load (count/mL fecal)",
       color = "Injury Status",
       title = "Parasite loads ranked - lowest to highest",
       subtitle = "Colored by injury status"
  )

```

The Wilcoxon test first sorts all parasite loads from the smallest to the largest and assigns each observation a rank. **If several gorillas have exactly the same parasite load, they tie**. Rather than breaking the tie arbitrarily, R gives each of them the **average of the ranks** they would have occupied



## 5. Kruskal-Wallis Test: Alternative to ANOVA

The Kruskal-Wallis test is often used when we want to compare three or more groups, but the response variable is not approximately normal.

### 5.1 Look at data
```{r}

# Question: Does parasite load differ among classes?
hist(data$parasite_load)

data <- data |> 
  mutate(age_class = factor(age_class, levels = c("infant", "subadult", "adult")))

ggplot(data, aes(x = age_class, y = parasite_load)) +
  geom_boxplot(outlier.color = "red") +
  theme_bw()

data |> 
  group_by(age_class) |> 
  summarize(
    n = n(),
    mean_parasites = mean(parasite_load, na.rm = TRUE),
    sd_parasites = sd(parasite_load, na.rm = TRUE)
  )

```


### 5.2 Run the Kruskal-Wallis Test

```{r}

kruskal.test(parasite_load ~ age_class, data)

```

The Kruskal-Wallis test asks:

> Do parasite load values tend to differ among age classes?

It does this by ranking all parasite load values from lowest to highest and then asking whether one age class tends to have consistently higher (or lower) ranks than the others.


### 5.2 Post-hoc test
The Kruskal-Wallis test tells us whether at least one group differs, but not which groups differ.

If the Kruskal-Wallis test is significant, we can perform Dunn's test, the non-parametric equivalent of Tukey's HSD, to compare every pair of groups while adjusting for multiple comparisons.

Just like Tukey adjusted the p-values after ANOVA, Dunn's test adjusts the p-values because we're making several pairwise comparisons.
```{r}

# from the rstatix package
dunn_test(data, parasite_load ~ age_class, p.adjust.method = "holm")

```



## 6. Count Data

Now we will move from nonparametric group comparisons to models designed specifically for count data.

A **Poisson** distribution describes the pattern we often see when counting how many times something happens. A Poisson distribution has several characteristics:

- The values are counts (0, 1, 2, 3, ...)
- Negative values are impossible.
- Smaller counts are usually more common than larger counts.
- The distribution is often right-skewed (many small values, few large values).
- A very important property of a Poisson distribution is that **the mean and variance are expected to be approximately equal**.

### 6.1 Summary
```{r}

summary(data$play_events)

mean(data$play_events)
var(data$play_events)

```

### Key idea

For a Poisson distribution, the mean and variance are expected to be similar.

If the variance is much larger than the mean, that may be evidence of **overdispersion**.



## 7. Why Not Just Use Linear Regression?

Let's ask:

> Do older gorillas have higher parasite loads?

A simple linear model might seem like an obvious option.

### 7.1 Visualize
```{r}

ggplot(data, aes(x = age_years, y = play_events)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", color = "red") +
  theme_bw()

mod_lm <- lm(play_events ~ age_years, data = data)

summary(mod_lm)
```



## 8. Poisson Regression

Poisson regression is a type of **generalized linear model**, or **GLM**. It is commonly used when the response variable is a count.
A GLM extends linear regression so that we can analyze these different types of response variables. Instead of changing what we're trying to do, a GLM changes **how the model describes the response variable**.

### 8.1 Generalized linear model
```{r}

mod_glm <- glm(play_events ~ age_years, data = data, family = "poisson")

summary(mod_glm)

```

Interpretation: 
```text
- Estimate -> 	      Direction and size of the relationship (on the log scale)
- p-value	 ->         Is there evidence of a relationship?
exp(coef())	  ->      Biological interpretation (percent change in expected count)
Residual deviance -> 	The model fits much better than the null model
AIC	->                Used only to compare competing models (e.g., Poisson vs. Negative Binomial)
```

### 8.1 Interpreting Poisson Coefficients

Poisson regression uses a log link. That means the coefficients are on the log scale. To make them easier to interpret, we exponentiate them using `exp()`.

```{r}

coef(mod_glm)
# > coef(mod_glm)
# (Intercept)   age_years 
#   3.3330633  -0.1605488 

exp(coef(mod_glm))
# > exp(coef(mod_glm))
# (Intercept)   age_years 
#  28.0240566   0.8516763 

```

> `exp()`, or `e^x` is the natural exponential function, and it's the inverse of the natural logarithm (`log()`).
In a **linear regression**, every additional year changes the response by **adding** or **subtracting** the same amount.  
In a **Poisson regression**, every additional year changes the response by **multiplying** by the same number. 


## 9. Predicted Values from the Poisson Model

Predicted values help us understand the model in the original units of the response variable.
Before we can draw the prediction line from our Poisson model, we need a set of ages at which we want to make predictions.

### 9.1 Wrangle
```{r}

# This code creates a new dataset called new_age that contains 
# 100 evenly spaced ages covering the full range of our data.

# Create new data for prediction
new_age <- tibble(
  age_years = seq(
    from = min(data$age_years, na.rm = TRUE),
    to = max(data$age_years, na.rm = TRUE),
    length.out = 100
  )
) # seq() creates a sequence of numbers between the two ages, min/max.

# Get predicted values on the response scale
new_age$predicted_play_events <- predict(
  mod_glm,
  newdata = new_age,
  type = "response"
) 
# Uses the fitted Poisson model (mod_glm) to make predictions.
# Remember: we're not fitting a new model here. We're using the model we already built.

# By default, Poisson regression predicts on the log scale, 
# because that's how the model is fitted.
# type = "response" converts those predictions back to the original scale of the data.

head(new_age)
tail(new_age)
```

### 9.2 Visualize
```{r}

ggplot(data, aes(x = age_years, y = play_events)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "lm") +
  geom_line(data = new_age, aes(x = age_years, y = predicted_play_events), color = "red", linewidth = 1) +
  theme_bw() +
  labs(x = "Age",
       y = "Play events",
       title = "GLM: predicted play by age",
       subtitle = "red = model\nblue = smoothed raw data"
       )

AIC(mod_lm, mod_glm)

```

This plot is useful because it shows the model prediction in the original units of the data.



## 10. Summary: Choosing a Model When Normality Fails

Ask:

1. What is the response variable?
2. Is it quantitative, count, binary, or proportion?
3. Does the response look approximately normal?
4. Are there many zeros?
5. Is the variance much larger than the mean?
6. Are we comparing groups or making predictions?
7. Which model matches the response variable?

A simplified guide:

| Situation | Possible analysis |
|---|---|
| Two groups, non-normal quantitative response | Wilcoxon test |
| Three or more groups, non-normal quantitative response | Kruskal-Wallis test |
| Count response, mean similar to variance | Poisson regression |
| Count response, variance much larger than mean | Negative binomial regression |
| Positive continuous response, right-skewed | Consider log transformation |
| Binary response | Logistic regression, next week |

# Skills Application -- Lab

## 11. Lab Challenge: Non-normal and Count Data

For this lab, you will work with your own dataset.

Your goal is to identify whether you have a response variable that is non-normal, skewed, or count-based, and choose an appropriate analysis.

### Task List

1. Load your dataset and continue working in your script.
2. Choose one response variable that is not approximately normal, or choose one count variable.
3. Inspect the response variable using `summary()` and a plot.
4. Decide what type of response variable it is:
   - skewed continuous
   - count
   - binary
   - proportion
5. If you are comparing groups:
   - consider `wilcox.test()` for two groups
   - consider `kruskal.test()` for three or more groups
6. If your response is a count:
   - fit a Poisson model with `glm(..., family = "poisson")`
   - check for overdispersion
   - if needed, fit a negative binomial model with `MASS::glm.nb()`
7. Write 2-3 sentences interpreting your results biologically.
