---
title: "Week 3: When Normality Is Not Appropriate"
author: "Ronnie Steinitz"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 200
---

```{r global_options, include=FALSE}
options(width = 60)
```

# Skills Learning -- Lecture

Last week, we focused on comparing groups when the response variable was quantitative and reasonably close to normal.

This week, we will ask:

> **What do we do when normality is not a good assumption?**

This happens often in ecological and biological data. Many response variables are not normally distributed.

Examples include:

- count data, such as aggression events or parasite load
- skewed data, such as hormone concentrations
- data with many zeros
- data that cannot be negative

Today we will focus on:

- recognizing when a t-test, ANOVA, or linear model may not be appropriate
- nonparametric alternatives to t-tests and ANOVA
- count data
- Poisson regression
- overdispersion
- negative binomial regression
- when log transformations may or may not help

The main question for today is:

> **What kind of analysis should I use when my response variable is not normally distributed?**

## 0. Load Required Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
library(MASS)
```

## 1. Load and Preview the Dataset

Today we will continue using the faux gorilla hormone + behavioral dataset.

```{r}
# load dataset
data <- readr::read_csv(here::here("Week 1/gorilla_hormone_data.csv")) %>% 
  janitor::clean_names()

# preview dataset
glimpse(data, width = 80)
```

## 2. When Normality Is Not Appropriate

Many standard tests assume that the response variable, or the model residuals, are approximately normal.

This is often reasonable for continuous variables like:

- body mass
- hormone concentration
- temperature
- age

But it is often not reasonable for variables like:

- number of aggression events
- parasite load
- number of nests
- number of offspring
- number of animal sightings

Count variables have special properties:

- they are whole numbers
- they cannot be negative
- they often have many zeros
- they are often right-skewed
- the variance often increases as the mean increases

```{r}
# Look at continuous and count variables
summary(data$cortisol_ng_ml)
summary(data$aggression_events)
summary(data$parasite_load)

# Count how many observations have each aggression value
data %>% count(aggression_events)

# Count how many observations have each parasite load value
data %>% count(parasite_load)
```

### 2.1 Compare Distributions

```{r}
# Continuous variable
ggplot(data, aes(x = cortisol_ng_ml)) +
  geom_histogram() +
  labs(
    x = "Cortisol (ng/mL)",
    y = "Count",
    title = "Continuous response: cortisol"
  )

# Count variable
ggplot(data, aes(x = aggression_events)) +
  geom_histogram(binwidth = 1) +
  labs(
    x = "Aggression events",
    y = "Count",
    title = "Count response: aggression events"
  )

# Another count variable
ggplot(data, aes(x = parasite_load)) +
  geom_histogram(binwidth = 5) +
  labs(
    x = "Parasite load",
    y = "Count",
    title = "Count response: parasite load"
  )
```

### Key idea

A model should match the type of response variable.

```text
Quantitative and approximately normal  ->  t-test, ANOVA, linear model
Non-normal group comparison             ->  Wilcoxon or Kruskal-Wallis
Count response                           ->  Poisson or negative binomial model
Binary response                          ->  logistic regression
Repeated measures                        ->  mixed effects model
```

This week we will focus on non-normal group comparisons and count responses.

## 3. Nonparametric Alternatives

Last week, we used t-tests and ANOVA to compare group means.

But what if the response variable is strongly skewed, has extreme outliers, or does not look approximately normal?

In those cases, we may use nonparametric tests.

Nonparametric tests are useful because they do not rely as strongly on the assumption that the response variable is normally distributed.

Today we will use:

- `wilcox.test()` as an alternative to a two-sample t-test
- `kruskal.test()` as an alternative to ANOVA

Important note:

> Nonparametric tests are not automatically better. They are used when the assumptions of the parametric test are not reasonable.

## 4. Wilcoxon Test: Alternative to a Two-Sample t-test

The Wilcoxon test is often used when we want to compare two independent groups, but the response variable is not approximately normal.

Example question:

> Do injured and uninjured gorillas differ in parasite load?

Response variable:

- `parasite_load`

Grouping variable:

- `injury_status`

Number of groups:

- 2

```{r}
# Make injury labels easier to read
data <- data %>%
  mutate(
    injury_label = if_else(injury_status == 1, "Injured", "Not injured")
  )

# Visualize parasite load by injury status
ggplot(data, aes(x = injury_label, y = parasite_load)) +
  geom_boxplot(outlier.colour = "red") +
  labs(
    x = "Injury status",
    y = "Parasite load",
    title = "Parasite load by injury status"
  )
```

```{r}
# Summary by group
data %>%
  group_by(injury_label) %>%
  summarise(
    n = n(),
    mean_parasites = mean(parasite_load, na.rm = TRUE),
    median_parasites = median(parasite_load, na.rm = TRUE),
    sd_parasites = sd(parasite_load, na.rm = TRUE),
    .groups = "drop"
  )
```

### 4.1 Run the Wilcoxon Test

```{r}
wilcox.test(parasite_load ~ injury_label, data = data)
```

### How to interpret this

The Wilcoxon test asks:

> Do the values in one group tend to be higher or lower than the values in the other group?

This is slightly different from a t-test, which focuses on the difference between group means.

For non-normal or skewed data, it is often useful to report the median as well as the mean.

## 5. Kruskal-Wallis Test: Alternative to ANOVA

The Kruskal-Wallis test is often used when we want to compare three or more groups, but the response variable is not approximately normal.

Example question:

> Does parasite load differ among age classes?

Response variable:

- `parasite_load`

Grouping variable:

- `age_class`

Number of groups:

- 3 or more

```{r}
# Put age classes in biological order
data <- data %>% 
  mutate(age_class = factor(age_class, levels = c("infant", "subadult", "adult")))

# Visualize parasite load by age class
ggplot(data, aes(x = age_class, y = parasite_load)) +
  geom_boxplot(outlier.colour = "red") +
  labs(
    x = "Age class",
    y = "Parasite load",
    title = "Parasite load by age class"
  )
```

```{r}
# Summary by age class
data %>%
  group_by(age_class) %>%
  summarise(
    n = n(),
    mean_parasites = mean(parasite_load, na.rm = TRUE),
    median_parasites = median(parasite_load, na.rm = TRUE),
    sd_parasites = sd(parasite_load, na.rm = TRUE),
    .groups = "drop"
  )
```

### 5.1 Run the Kruskal-Wallis Test

```{r}
kruskal.test(parasite_load ~ age_class, data = data)
```

### How to interpret this

The Kruskal-Wallis test asks:

> Do parasite load values tend to differ among age classes?

It is similar in purpose to ANOVA, but it is used when ANOVA assumptions are not reasonable.

Like ANOVA, Kruskal-Wallis tells us whether there is evidence of a difference somewhere among the groups. It does not tell us exactly which groups differ.

## 6. Count Data

Now we will move from nonparametric group comparisons to models designed specifically for count data.

Count data are common in ecology and wildlife research.

Examples:

- number of aggression events
- number of parasites
- number of nests
- number of vocalizations
- number of offspring
- number of animal sightings

Count data are different from continuous data because counts:

- are whole numbers
- cannot be negative
- often include zeros
- often have variance that increases with the mean

```{r}
# Inspect count variables
summary(data$aggression_events)
summary(data$parasite_load)

# Mean and variance of aggression events
mean(data$aggression_events)
var(data$aggression_events)

# Mean and variance of parasite load
mean(data$parasite_load)
var(data$parasite_load)
```

### Key idea

For a Poisson distribution, the mean and variance are expected to be similar.

If the variance is much larger than the mean, that may be evidence of overdispersion.

## 7. Why Not Just Use Linear Regression?

Let's ask:

> Do older gorillas have higher parasite loads?

A simple linear model might seem like an obvious option.

```{r}
# Linear model
mod_lm <- lm(parasite_load ~ age_years, data = data)

summary(mod_lm)
```

But linear regression can be inappropriate for count data because it can predict impossible values, including negative counts.

```{r}
# Predicted values from linear model
data_lm <- data %>%
  mutate(pred_lm = predict(mod_lm))

summary(data_lm$pred_lm)

# Check if any predicted values are negative
data_lm %>%
  filter(pred_lm < 0) %>%
  select(sample_id, gorilla_id, age_years, parasite_load, pred_lm)
```

Even if this particular model does not predict negative values, a linear model is still not ideal for count data because it assumes normally distributed errors and constant variance.

## 8. Poisson Regression

Poisson regression is a type of generalized linear model, or GLM.

It is commonly used when the response variable is a count.

In R, we use:

```text
glm(..., family = "poisson")
```

Example question:

> Does parasite load change with age?

Response variable:

- `parasite_load`, a count

Predictor variable:

- `age_years`, continuous

```{r}
# Poisson regression
mod_pois <- glm(
  parasite_load ~ age_years,
  data = data,
  family = "poisson"
)

summary(mod_pois)
```

### 8.1 Interpreting Poisson Coefficients

Poisson regression uses a log link.

That means the coefficients are on the log scale.

To make them easier to interpret, we exponentiate them using `exp()`.

```{r}
# Coefficients on the log scale
coef(mod_pois)

# Exponentiated coefficients
exp(coef(mod_pois))
```

The exponentiated coefficient tells us the multiplicative change in the expected count for a one-unit increase in the predictor.

For example:

> If `exp(coef)` for age is 1.03, then each additional year of age is associated with a 3% increase in expected parasite load.

```{r}
# Make a clean coefficient table
poisson_results <- tibble(
  term = names(coef(mod_pois)),
  estimate_log_scale = coef(mod_pois),
  estimate_count_ratio = exp(coef(mod_pois))
)

poisson_results
```

## 9. Predicted Values from the Poisson Model

Predicted values help us understand the model in the original units of the response variable.

```{r}
# Create new data for prediction
new_age <- tibble(
  age_years = seq(
    from = min(data$age_years, na.rm = TRUE),
    to = max(data$age_years, na.rm = TRUE),
    length.out = 100
  )
)

# Get predicted values on the response scale
new_age$predicted_parasite_load <- predict(
  mod_pois,
  newdata = new_age,
  type = "response"
)

head(new_age)
```

```{r}
# Plot observed data and predicted values
ggplot(data, aes(x = age_years, y = parasite_load)) +
  geom_point(alpha = 0.4) +
  geom_line(
    data = new_age,
    aes(x = age_years, y = predicted_parasite_load),
    color = "red",
    linewidth = 1
  ) +
  labs(
    x = "Age (years)",
    y = "Parasite load",
    title = "Poisson model: predicted parasite load by age"
  )
```

This plot is useful because it shows the model prediction in the original units of the data.

## 10. Overdispersion

Poisson regression assumes that the mean and variance are similar.

Ecological data often violate this assumption.

When the variance is much larger than the mean, we call this **overdispersion**.

Overdispersion matters because it can make p-values too small and make us too confident in the results.

A simple way to check for overdispersion is:

```text
residual deviance / residual degrees of freedom
```

Values close to 1 are usually acceptable.

Values much larger than 1 suggest overdispersion.

```{r}
# Check overdispersion
dispersion_ratio <- deviance(mod_pois) / df.residual(mod_pois)

dispersion_ratio
```

### How to interpret this

- close to 1: little evidence of overdispersion
- greater than 1.5 or 2: possible overdispersion
- much larger than 2: strong evidence of overdispersion

This is a rule of thumb, not a strict law.

## 11. Negative Binomial Regression

Negative binomial regression is useful when the response variable is a count and the data are overdispersed.

A negative binomial model is like a Poisson model, but it allows extra variation.

In R, we can use `glm.nb()` from the `MASS` package.

```{r}
# Negative binomial regression
mod_nb <- MASS::glm.nb(
  parasite_load ~ age_years,
  data = data
)

summary(mod_nb)
```

### 11.1 Compare Poisson and Negative Binomial Models

One simple way to compare models is AIC.

Lower AIC values suggest better model fit, relative to the other model being compared.

```{r}
AIC(mod_pois, mod_nb)
```

If the negative binomial model has a much lower AIC, it may fit the count data better than the Poisson model.

### 11.2 Predicted Values from the Negative Binomial Model

```{r}
new_age$predicted_nb <- predict(
  mod_nb,
  newdata = new_age,
  type = "response"
)

head(new_age)
```

```{r}
ggplot(data, aes(x = age_years, y = parasite_load)) +
  geom_point(alpha = 0.4) +
  geom_line(
    data = new_age,
    aes(x = age_years, y = predicted_nb),
    color = "blue",
    linewidth = 1
  ) +
  labs(
    x = "Age (years)",
    y = "Parasite load",
    title = "Negative binomial model: predicted parasite load by age"
  )
```

## 12. When Should We Log-Transform?

Log transformations can help when a continuous response variable is strongly right-skewed.

They are often used for positive continuous variables such as:

- hormone concentrations
- body mass
- home range size
- concentration values

```{r}
# Cortisol distribution
ggplot(data, aes(x = cortisol_ng_ml)) +
  geom_histogram() +
  labs(
    x = "Cortisol (ng/mL)",
    y = "Count",
    title = "Raw cortisol values"
  )

# Log cortisol distribution
ggplot(data, aes(x = log(cortisol_ng_ml))) +
  geom_histogram() +
  labs(
    x = "log(Cortisol)",
    y = "Count",
    title = "Log-transformed cortisol values"
  )
```

### Important cautions

A log transformation is not always appropriate.

Be careful when:

- the response variable is a count
- the response variable includes zeros
- the response variable is binary
- the response variable is a proportion bounded between 0 and 1

For count data, Poisson or negative binomial models are usually more appropriate than simply logging the count and using a linear model.

## 13. Summary: Choosing a Model When Normality Fails

Ask:

1. What is the response variable?
2. Is it quantitative, count, binary, or proportion?
3. Does the response look approximately normal?
4. Are there many zeros?
5. Is the variance much larger than the mean?
6. Are we comparing groups or making predictions?
7. Which model matches the response variable?

A simplified guide:

| Situation | Possible analysis |
|---|---|
| Two groups, non-normal quantitative response | Wilcoxon test |
| Three or more groups, non-normal quantitative response | Kruskal-Wallis test |
| Count response, mean similar to variance | Poisson regression |
| Count response, variance much larger than mean | Negative binomial regression |
| Positive continuous response, right-skewed | Consider log transformation |
| Binary response | Logistic regression, next week |

# Skills Application -- Lab

## 14. Lab Challenge: Non-normal and Count Data

For this lab, you will work with your own dataset.

Your goal is to identify whether you have a response variable that is non-normal, skewed, or count-based, and choose an appropriate analysis.

### Task List

1. Load your dataset and continue working in your script.
2. Choose one response variable that is not approximately normal, or choose one count variable.
3. Inspect the response variable using `summary()` and a plot.
4. Decide what type of response variable it is:
   - skewed continuous
   - count
   - binary
   - proportion
5. If you are comparing groups:
   - consider `wilcox.test()` for two groups
   - consider `kruskal.test()` for three or more groups
6. If your response is a count:
   - fit a Poisson model with `glm(..., family = "poisson")`
   - check for overdispersion
   - if needed, fit a negative binomial model with `MASS::glm.nb()`
7. Write 2-3 sentences interpreting your results biologically.

## 15. Template for Your Answer

**Biological question:**  

**Response variable:**  

**Response variable type:**  

**Why might a normal model not be appropriate?**  

**Predictor or grouping variable:**  

**Chosen analysis:**  

**Why this analysis?**  

**Biological interpretation:**  

```{r}
# Load your data here
```

```{r}
# Inspect your response variable here
```

```{r}
# Run your analysis here
```

```{r}
# Interpret or visualize results here, if useful
```
