---
title: "Week 1: From Research Question to Statistical Model"
author: "Author"
date: "`r Sys.Date()`"
output:
  html_document: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 200
---

# Skills Learning -- Lecture

This week, we are shifting from basic R skills toward statistical thinking.

In the introductory course, we focused on importing data, exploring data, cleaning columns, summarizing variables, and making plots. In this advanced course, we will build on those skills and ask a new question:

> **Given my biological question and my data, what kind of statistical model should I use?**


## 0. Load Required Packages
```{r, message=FALSE, warning=FALSE}

# Load packages
library(tidyverse)
library(janitor)
library(here)

```


## 1. Load and Preview the Dataset
```{r}

# Load dataset
data <- readr::read_csv(here::here("Week 1/gorilla_hormone_data.csv")) %>% 
  janitor::clean_names()

glimpse(data)

```

## 2. Numeric Variables
```{r}

summary(data$cortisol_ng_ml)
summary(data$sex)
summary(data$parasite_load)

class(data$cortisol_ng_ml)
class(data$parasite_load)

mean_cort <- mean(data$cortisol_ng_ml)

ggplot(data, aes(x = cortisol_ng_ml)) +
  geom_histogram() +
  geom_vline(aes(xintercept = mean_cort),
             color = "red",
             size = 1,
             linetype = "dashed"
             )
# "dashed", "dotted", "solid", ...

```

### 2.2 Count Variables
```{r}

summary(data$aggression_events)

ggplot(data, aes(x = aggression_events)) +
  geom_histogram()

```

### 2.3 Binary Variables
```{r}

class(data$respiratory_signs)
summary(data$respiratory_signs)

data %>% 
  count(respiratory_signs)

mean(data$respiratory_signs)

data %>% 
  count(injury_status)

prop_injured <- 20/240
prop_injured

mean(data$injury_status)

ggplot(data, aes(x = respiratory_signs)) +
  geom_histogram()

```


### 2.4 Proportion Variables
```{r}

class(data$feeding_time_prop)
summary(data$feeding_time_prop)

ggplot(data, aes(x = feeding_time_prop)) +
  geom_histogram()

```


## 3. ANALYSIS
```{r}

ggplot(data, aes(x = aggression_events, y = cortisol_ng_ml)) +
  geom_point() +
  geom_smooth(method = "lm")

mod1 <- lm(data = data, cortisol_ng_ml ~ aggression_events)
summary(mod1)


mod2 <- lm(data = data, cortisol_ng_ml ~ aggression_events + injury_status)

summary(mod2)

ggplot(data, aes(x = aggression_events, y = cortisol_ng_ml, group = injury_status)) +
  geom_point(aes(color = as.factor(injury_status))) +
  geom_smooth(method = "lm", aes(color = as.factor(injury_status)))

```


## 4. Considerations
### 4.1 Important model-selection questions

Before choosing a model, ask:

1. What is the response variable?
2. What type of variable is the response?
3. Is the response approximately normal, skewed, count, binary, or proportion?
4. Are observations independent?
5. Are individuals sampled repeatedly?
6. Is the dataset cross-sectional or longitudinal?
7. What are the predictor variables?
8. Are predictors continuous, categorical, binary, count, or proportion?
9. Does the research question involve group differences, relationships, prediction, or repeated measures?


# Skills Application -- Lab

## 5. Lab Challenge: Choose the Model Type

For this lab, your goal is **not** to run the final model yet.

Your goal is to look at a biological question, identify the response variable, diagnose the structure of the data, and decide what kind of model might be appropriate.

For each question below, answer:

1. What is the response variable?
2. What type of variable is the response?
3. What are the predictor variables?
4. Are the observations independent?
5. Is the dataset cross-sectional or longitudinal/repeated?
6. What plot would help you inspect the response?
7. What general model type might be appropriate?


### 5.1 Question 1

> Do gorillas with more aggression events have higher cortisol?

```{r}

# Make at least one plot to inspect this question here

```

**Your answer:**  
_Response variable:_  
_Response type:_  
_Predictor variable(s):_  
_Independent or repeated observations?_  
_Cross-sectional or longitudinal?_  
_Useful plot:_  
_Possible model type:_  


### 5.2 Question 2

> Are respiratory signs more likely in larger groups?

```{r}

# Make at least one plot to inspect this question here

```

**Your answer:**  
_Response variable:_  
_Response type:_  
_Predictor variable(s):_  
_Independent or repeated observations?_  
_Cross-sectional or longitudinal?_  
_Useful plot:_  
_Possible model type:_  


### 5.3 Question 3

> Does feeding time proportion differ by group?

```{r}

# Make at least one plot to inspect this question here

```

**Your answer:**  
_Response variable:_  
_Response type:_  
_Predictor variable(s):_  
_Independent or repeated observations?_  
_Cross-sectional or longitudinal?_  
_Useful plot:_  
_Possible model type:_  


### 5.4 Optional: Your own biological question

Write one biological question using this dataset or your own dataset.

**Your question:**  

```{r}

# Make at least one plot to inspect your question here

```

**Your answer:**  
_Response variable:_  
_Response type:_  
_Predictor variable(s):_  
_Independent or repeated observations?_  
_Cross-sectional or longitudinal?_  
_Useful plot:_  
_Possible model type:_  

------------------------------------------------------------------------

