---
title: 'Week 2: Comparing Groups'
author: "Ronnie Steinitz"
date: "`r Sys.Date()`"
output:
  html_document: default
editor_options:
  chunk_output_type: console
  markdown:
    wrap: 200
---

# Skills Learning -- Lecture

Last week, we focused on identifying different types of variables and thinking about how the structure of our data affects the type of analysis we can use.

This week, we will focus on one major branch of the statistical decision tree:

> **Comparing groups when the response variable is quantitative.**

Today we will focus on:

- one-sample t-tests
- independent two-sample t-tests
- paired t-tests
- ANOVA
- visualizing group comparisons
- interpreting results biologically

The main question for today is:

> **I have a quantitative response variable and I want to compare groups. Which test should I use?**

## 0. Load Required Packages

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(here)
```

## 1. Load and Preview the Dataset

Today we will continue using the faux gorilla hormone + behavioral dataset.

Each row represents one sample or observation from a gorilla. Some gorillas appear more than once, which means the dataset contains repeated measures. Later in the course, we will learn how to model repeated measures directly. Today, we will focus on the logic of comparing groups.

```{r}

# load dataset
data <- readr::read_csv(here::here("Week 1/gorilla_hormone_data.csv")) %>% 
  janitor::clean_names()

# preview dataset
glimpse(data, width = 80)

```

## 2. What Does "Comparing Groups" Mean?

A group comparison asks whether a quantitative response variable differs across categories.


### 2.1 summary stats
```{r}


```

## 3. One-Sample t-test

A one-sample t-test compares the mean of one group to a known or expected value.

### 3.1 Test
```{r}


```



## 4. Visualizing Group Comparisons
### 4.1 Boxplot

```{r}


```


## 5. Independent Two-Sample t-test

An independent two-sample t-test compares the mean of a quantitative variable between two independent groups.

### 5.1 Independent t-test
```{r}


```


### 5.2 Plots

```{r}


```




## 6. Paired t-test

A paired t-test is used when each observation in one group has a natural partner in the other group.

### 6.1 Import paired data

```{r}

paired_data <- read_csv(here::here("Week 2/gorilla_paired.csv")) %>% 
  janitor::clean_names()

```

### 6.2 Run t-test
```{r}


```

### 6.3 Plot
```{r}


```


## 7. ANOVA: Comparing 3 or More Groups

ANOVA is used when we want to compare the mean of a quantitative response variable across three or more groups.

### 7.1 Plot
```{r}


```

### 7.2 ANOVA
```{r}


```

### 7.3 Tukey 
HSD = Honestly Significant Difference
```{r}


```

- The `diff` column gives the estimated difference between the two group means.
- The `lwr` and `upr` columns give the 95% confidence interval for that difference.
- The `p adj` column contains the *adjusted* p-value, which accounts for the fact that multiple pairwise comparisons are being made.

> You should only run a Tukey test if you first have a reason to ask which groups differ; i.e., if the ANOVA came back significant. 


## 8. Checking Assumptions

For t-tests and ANOVA, we often want to know whether the response variable is approximately normally distributed within groups.

### 8.1 Histograms

```{r}


```

### 8.2 QQ Plot

A QQ plot helps us check whether the data are approximately normal.

```{r}


```

### 8.3 Model Diagnostic Plots

```{r}


```



## 9. Summary: Choosing the Right Group Comparison

When comparing groups with a quantitative response variable, ask:

1. What is the response variable?
2. Is the response variable quantitative?
3. What is the grouping variable?
4. How many groups are there?
5. Are the groups independent or paired?
6. Do the data look approximately normal?
7. Which test matches the question and data structure?


# Skills Application -- Lab

## 10. Lab Challenge: Compare Groups in Your Own Dataset

For this lab, you will work with your own dataset.

Your goal is to choose a quantitative response variable and compare it across groups.

You do not need to run every test. Focus on choosing the correct test for your question.

### Task List

1. Load your dataset and continue working in your script
2. Choose one quantitative response variable
3. Choose one grouping variable; Count how many groups are in your grouping variable
4. Create a plot comparing the response variable across groups
5. Suggest a biological question you can answer using these variables
6. Decide which test is most appropriate and run a test
7. Write 2-3 sentences interpreting the results, biologically

> Goal: perform an analysis from start to finish!
Research question --> Data --> Analysis --> Biological Insight
